Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2013-10-06 paper/paper.tex
Move C- section to top. Add explanation of DFD diagrams

Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
which is the closest work to ours.  {\sc Brahma} also focuses on
synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
authors characterise the program synthesis problem as a second-order
$\exists \forall$ formula, and introduce the CEGIS algorithm for solving
these formulae.  This algorithm reduces the satisfiability of the $\exists
\forall$ formula to the satisfiability of several first-order formulae, each
of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
takes as input a specification (given as an SMT formula) and a library of
components (one component is more or less a single machine instruction).  It
then finds a permutation of the component library that satisfies the
specification.  This permutation is the final synthesised program.  In the
event that the library does not contain enough components to synthesise a
correct program, the user is told that the library is too small and is
required to add more components until the synthesis succeeds.  Each round of
synthesis amounts to solving an $\exists \forall$ formula that is quadratic
in the size of the component library.

We require the user to supply a program specification written in C, but then
ask nothing more of her.  Our encoding allows us to generate an $\exists
\forall$ formula that is \emph{linear} in the length of the shortest program
satisfying the user's specification, which we then check for satisfiability
using the CEGIS algorithm.  In contrast, the program generated by {\sc
brahma} can be no longer than the size of the library, as the synthesis
outcome is a permutation of the component library.

Another successful approach to program synthesis has been \emph{program
sketching}, in particular the {\sc sketch} tool~\cite{sketch}.  This method
requires the programmer to provide a \emph{program sketch}, written in a
C-like language, together with either a specification (written in the same
language) or a suite of unit tests.  The sketch is a skeleton program which
contains much of the high-level structure of the final program, but with
several ``holes'', which {\sc sketch} fills in to meet the specification,
again using the CEGIS algorithm.

@@

\subsection{The Abstract Algorithm}

When the first-order fragment of the logic is efficiently decidable and $P$ can be
expressed as a ground term of the logic, we can use a refinement loop to check the
validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
and {\sc synth} synthesizes a candidate program which is correct on just those inputs.
{\sc Verif} then tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
procedure is guaranteed to terminate.

To synthesize tricky bitvector programs like this, it is natural for us to
work in the logic of quantifier-free propositional formulae and to use a
propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
However, we propose a slightly different tack, which is to use a decidable
fragment of C as our underlying logic.  The fragment we use has the
following restrictions:

\begin{itemize}
 \item All loops and recursive function calls must terminate after a constant number
 of iterations.  This constant must be statically inferrable.
 \item Arrays must all be statically allocated with a constant size.
\end{itemize}

Any other C constructs are permitted, which still leaves a very expressive
language with the nice property that safety is decidable using a single
query to a bounded model checker.  We will refer to this fragment as \newC. 
To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
\sigma$ and $P$ in \newC, then ensure that we can express the validity of the
synthesis formula as a safety property of the resulting \newC program.

\begin{figure}
\centering
\begin{minipage}{0.65\textwidth}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

\begin{tabular}{lll}
 \verb|add a b| & \verb|sub a b| & \verb|mul a b| \\
 \verb|div a b| & \verb|neg a|
\end{tabular}
\bigskip

\begin{tabular}{lll}
 \verb|and  a b| & \verb|or   a b| & \verb|xor  a b| \\
 \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
\end{tabular}
\bigskip

\begin{tabular}{lll}
 \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
 \verb|slt a b|
\end{tabular}
\end{minipage}

@@

We use a fragment of C as our underlying logic.  We will refer to this fragment
as \newC.  The characteristic property of \newC is that safety can be decided for
\newC programs by using a bounded model checker.  A \newC program is just a C program
with the following syntactic restrictions:

@@

>>>>

Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
which is the closest work to ours.  {\sc Brahma} also focuses on
synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
authors characterise the program synthesis problem as a second-order
$\exists \forall$ formula and uses the CEGIS algorithm,
introduced in~\cite{lezama-thesis}, for solving
these formulae.  This algorithm reduces the satisfiability of the $\exists
\forall$ formula to the satisfiability of several first-order formulae, each
of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
takes as input a specification (given as an SMT formula) and a library of
components (one component is more or less a single machine instruction).  It
then finds a permutation of the component library that satisfies the
specification.  This permutation is the final synthesised program.  In the
event that the library does not contain enough components to synthesise a
correct program, the user is told that the library is too small and is
must add more components until the synthesis succeeds.  Each round of
synthesis amounts to solving an $\exists \forall$ formula that is quadratic
in the size of the component library.

We require the user to supply a program specification written in C, but then
ask nothing more of her.  We parameterise the space of programs in such a way
that it can be explored automatically, rather than asking a human for hints.
Our program encoding allows generates an $\exists
\forall$ formula that is \emph{linear} in the length of the shortest program
satisfying the user's specification, which we then check for satisfiability
using the CEGIS algorithm.  In contrast, the program generated by {\sc
brahma} can be no longer than the size of the library, as the synthesis
outcome is a permutation of the component library.  Our synthesis
formulae are therefore asymptotically smaller than {\sc brahma}'s.

Another successful approach to program synthesis has been \emph{program
sketching}, in particular the {\sc sketch} tool~\cite{lezama-thesis,sketch}.  This method
requires the programmer to provide a \emph{program sketch}, written in a
C-like language, together with either a specification (written in the same
language) or a suite of unit tests.  The sketch is a skeleton program which
contains much of the high-level structure of the final program, but with
several ``holes'', which {\sc sketch} fills in to meet the specification,
again using the CEGIS algorithm.

@@

@@

We use a fragment of C as our underlying logic.  We will refer to this fragment
as \newC.  The characteristic property of \newC is that safety can be decided for
\newC programs using a single query to bounded model checker.  A \newC program is
just a C program with the following syntactic restrictions:

@@

\subsection{The Synthesis Formula}

\subsection{CEGIS}

When the first-order fragment of the logic is efficiently decidable and $P$ can be
expressed as a ground term of the logic, we can use a refinement loop to check the
validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
and {\sc synth} synthesizes a candidate program that is correct on just those inputs.
{\sc Verif} then tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
procedure is guaranteed to terminate.

To synthesize tricky bitvector programs like this, it is natural for us to
work in the logic of quantifier-free propositional formulae and to use a
propositional SAT or SMT-$\mathcal{BV}$ solver as the decision procedure. 
However, we propose a slightly different tack, which is to use \newC as
our underlying logic.

To instantiate the abstract synthesis algorithm in \newC we must express $I, O,
\sigma$ and $P$ in \newC, then ensure that we can express the validity of the
synthesis formula as a safety property of the resulting \newC program.

\begin{figure}
\begin{center}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

\begin{tabular}{lllll}
 \verb|add a b| & \verb|sub a b| & \verb|mul a b| & \verb|div a b| & \verb|neg a|
\end{tabular}

\medskip

\begin{tabular}{llllll}
 \verb|and  a b| & \verb|or a b| & \verb|xor a b| & \verb|ashr a b| & \verb|lshr a b| & \verb|not  a|
\end{tabular}

\medskip

\begin{tabular}{llll}
 \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| & \verb|slt a b|
\end{tabular}
\end{center}

In order to determine the validity of the {\sc synth} formula, we can now
check the {\sc synth} program for safety.  The {\sc synth} program is a
\newC program, which means we can check its safety by invoking a bounded
model checker, such as {\sc cbmc}.  Since \newC is a fragment of C,
the {\sc synth} program can also be compiled executable using {\sc gcc}.
We can add a small piece of code which searches all possible
$\mathcal{L}$-programs and evaluates each on the current test vectors until
it finds one which is correct, just like a conventional superoptimiser,
and compile the whole thing down to an executable.  This leaves
us with a very simple explicit-state model checker which decides
safety for {\sc synth}.  We can run the explicit-state model checker
in parallel with {\sc cbmc} and take the answer of whichever happens
to terminate first, stopping the other process.  This procedure is
depicted for {\sc synth} in Fig.~\ref{fig:synth-dfd} and for
{\sc verif} in Fig.~\ref{fig:verif-dfd}.
